---
title: "Data 605 HW9"
author: "William Jasmine"
date: "2023-03-20"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  pdf_document: default
---

# Question 1

## Description

This question comes from page 363, question #11 of ["Introduction to Probability"](https://math.dartmouth.edu/~prob/prob/prob.pdf) by Charles M. Grinstead.

The price of one share of stock in the Pilsdorff Beer Company (see Exercise 8.2.12) is given by $Y_n$ on the $n$th day of the year. Finn observes that
the differences $X_n = Y_{n+1} − Y_n$ appear to be independent random variables
with a common distribution having mean $\mu = 0$ and variance $\sigma^2 = 1/4$. If $Y_1$ = 100, estimate the probability that $Y_{365}$ is

(a) $\geq 100$.
(b) $\geq 110$.
(c) $\geq 120$.

## Solution
### Simulation

We know that $Y_{365} = Y_1 + X_1 + X_2 + ... + X_{364} = Y_1 + \sum_{i=1}^{364}X_i$. Given that $\sum_{i=1}^{364}X_i$ is a sum of independent random variables, we can assume it follows a normal distribution via the Central Limit Theorem. Furthermore, we can define the mean $\mu_{364}$ and standard deviation $\sigma_{364}$ of this normal distribution as: 

$$
\begin{align}
\mu_{364} &= 364\mu = 0 \\
\sigma_{364} &= \sqrt{\sigma^2_{X_1}+ \sigma^2_{X_1} + ... + \sigma^2_{364}} = \sqrt{\frac{364}{4}}
\end{align}
$$

We can use this mean and standard deviation to sample from a normal distribution and simulate 100,000 different values of $Y_{365}$:

```{r}
sd = sqrt(364/4)
sims = 100000

Y_365s = 100 + rnorm(sims, mean=0, sd=sd)
```

We can now use these simulated $Y_{365}$ values to estimate the probabilities:

```{r}
P_a <- length(which(Y_365s > 100)) / sims 
P_b <- length(which(Y_365s > 110)) / sims 
P_c <- length(which(Y_365s > 120)) / sims

str1 <- paste('Estimated probability of Y_365 >= 100:', P_a)
str2 <- paste('Estimated probability of Y_365 >= 110:', P_b)
str3 <- paste('Estimated probability of Y_365 >= 120:', P_c)
str_final <- paste(str1, str2, str3, sep = '\n')
writeLines(str_final)
```

```{r, eval=FALSE}
# the below can be used to check the above values
# using z-score and pnorm distribution

1-pnorm((100-100)/(0.5*sqrt(364)))
1-pnorm((110-100)/(0.5*sqrt(364)))
1-pnorm((120-100)/(0.5*sqrt(364)))
```

# Question 2

## Description
Calculate the expected value and variance of the binomial distribution using the moment generating function.

## Solution

Given a discrete random variable $X$ with probability mass function $f_X(x)$ and dummy parameter $t$, the moment generating function of $X$ is defined as the expectation value of $e^{tX}$:

$$
\begin{equation}
M_X(t) = E[e^{tX}]
\end{equation}
$$

Using the fact that, for a discrete random variable, $E[X] = \sum_{x \in X} x \cdot p(x)$, we have:

$$
\begin{equation}
M_X(t) = \sum_{x\in X}e^{tx} \cdot f_X(x)
\end{equation}
$$

Since $X$ follows a binomial distribution in which $0 \leq x \leq n$, we can write:

$$
\begin{align}
M_X(t) &= \sum_{x=0}^ne^{tx} \cdot {n \choose x} p^x q^{x-n} \\
M_X(t) &= \sum_{x=0}^n {n \choose x}(pe^t)^xq^{x-n}
\end{align}
$$

The binomial theorem tells us that $(u+v)^n = \sum_{x=0}^n {n \choose x}u^xv^{n-x}$, and as such we can simplify the above sum:

$$
\begin{equation}
M_X(t) = (pe^t + q)^n
\end{equation}
$$

Once we have the moment generating function of $X$, we can use it to determine $E[X]$ and $E[X^2]$:

$$
\begin{align}
E[X] &= M'_X(0) \\
E[X^2] &= M''_X(0)
\end{align}
$$

Thus, we first find $M'_X(t)$:

$$
\begin{align}
M'_X(t) &= \frac{d}{dt}((pe^t + q)^n) \\
M'_X(t) &= npe^t(pe^t+q)^{n-1} \\
\end{align}
$$

Using the above in conjunction with the fact that $p+q=1$, we can determine $E[X]$:

$$
\begin{align}
E[X] &= M'_X(0) \\
E[X] &= npe^0(pe^0+q)^{n-1} \\
E[X] &= np(p+q)^{n-1} \\
E[X] &= np(1)^{n-1} \\
E[X] &= np
\end{align}
$$

Next, we determine $M''_X(t)$:

$$
\begin{align}
M''_X(t) &= \frac{d}{dt}(npe^t(pe^t+q)^{n-1}) \\
M''_X(t) &= n(n-1)(pe^t)^2+npe^t(pe^t+q)^{n-2} \\
\end{align}
$$

Once again we evaluate at 0 to find $E[X^2]$:

$$
\begin{align}
E[X^2] &= M''_X(0) \\
E[X^2] &= n(n-1)(pe^0)^2+npe^0(pe^0+q)^{n-2} \\
E[X^2] &= n(n-1)(p^2)+np(p+q)^{n-2} \\
E[X^2] &= n^2p^2-np^2+ np\\
\end{align}
$$

To find the variance $\sigma^2$, we use the fact that $\sigma^2 = E[X^2] - E[X]^2$:

$$
\begin{align}
\sigma^2 &= n^2p^2-np^2+ np -(np)^2 \\
\sigma^2 &= np -np^2 \\
\sigma^2 &= np(1-p) \\
\sigma^2 &= npq
\end{align}
$$

Thus, to summarize, we have confirmed using the binomial distribution's moment generating function that: $E[X] = \mu = np$ and $\sigma^2 = npq$.

# Question 3

## Description 

Calculate the expected value and variance of the exponential distribution using the moment generating function.

## Solution 

Given a continuous random variable $X$ with probability density function $f_X(x)$ and dummy parameter $t$, the moment generating function of $X$ is defined as the expectation value of $e^{tX}$:

$$
\begin{equation}
M_X(t) = E[e^{tX}]
\end{equation}
$$

Using the fact that, for a continuous random variable, $E[X] = \int_{-\infty}^{\infty} x \cdot p(x)$, we have:

$$
\begin{equation}
M_X(t) = \int_{-\infty}^{\infty}e^{tx} \cdot f_X(x)\,dx
\end{equation}
$$

Since $X$ follows an exponential distribution in which:

$$
\begin{equation}
f_X(x) =
  \begin{cases}
        \lambda e^{-\lambda x} & \text{if } 0 \leq x < \infty\\
        0 & \text{otherwise }
  \end{cases}
\end{equation}
$$
we can write:

$$
\begin{align}
M_X(t) &= \int_{0}^{\infty}e^{tx} \lambda e^{-\lambda x} \, dx \\
M_X(t) &= \lambda \int_{0}^{\infty}e^{x(t-\lambda)} \, dx \\
M_X(t) &= \frac{\lambda}{t-\lambda}e^{x(t-\lambda)}|_0^{\infty}
\end{align}
$$

For $t<\lambda$, this integral evaluates to:

$$
\begin{align}
M_X(t) &= \frac{\lambda}{t-\lambda}(0-1) \\
M_X(t) &= \frac{\lambda}{\lambda - t}
\end{align}
$$

Now that we have the moment generating function of $X$, we can use it to determine $E[X]$ and $E[X^2]$:

$$
\begin{align}
E[X] &= M'_X(0) \\
E[X^2] &= M''_X(0)
\end{align}
$$

We first find $M'_X(t)$ below:

$$
\begin{align}
M'_X(t) &= \frac{d}{dt}\left( \frac{\lambda}{\lambda - t} \right) \\
M'_X(t) &= \frac{\lambda}{(\lambda-t)^2} \\
\end{align}
$$

Using the above, we can determine $E[X]$:

$$
\begin{align}
E[X] &= M'_X(0) \\
E[X] &= \frac{\lambda}{(\lambda-0)^2} \\
E[X] &= \frac{\lambda}{\lambda^2} \\
E[X] &= \frac{1}{\lambda} \\
\end{align}
$$

Next, we determine $M''_X(t)$:

$$
\begin{align}
M''_X(t) &= \frac{d}{dt}(\frac{\lambda}{(\lambda-t)^2}) \\
M''_X(t) &= \frac{2\lambda}{(\lambda-t)^3} \\
\end{align}
$$

Once again we evaluate at 0 to find $E[X^2]$:

$$
\begin{align}
E[X^2] &= M''_X(0) \\
E[X^2] &= \frac{2\lambda}{(\lambda-0)^3} \\
E[X^2] &= \frac{2}{\lambda^2} 
\end{align}
$$

To find the variance $\sigma^2$, we use the fact that $\sigma^2 = E[X^2] - E[X]^2$:

$$
\begin{align}
\sigma^2 &= \frac{2}{\lambda^2} - \left(\frac{1}{\lambda} \right)^2 \\
\sigma^2 &= \frac{1}{\lambda^2} 
\end{align}
$$

Thus, to summarize, we have confirmed using the exponential distribution's moment generating function that: $E[X] = \mu = \frac{1}{\lambda}$ and $\sigma^2 = \frac{1}{\lambda^2}$.



