---
title: "Data 607 - Extra Credit - Window Functions"
author: "William Jasmine"
date: "2022-09-14"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(openintro)
options(gsubfn.engine = "R")
library(sqldf)
```

The chunk below downloads the time series data from the "OpenIntro" package. More specifically, the data imported below is from the `sp500_1950_2018` dataset and includes daily financial metrics for the S&P 500 market for all trading days from 1950-2018. 
```{r}
data(sp500_1950_2018)
```

The query below demonstrates how to calculate both the the year-to-date average and the six-day moving averages for the opening and high daily price of the S&P 500 market. For simplicity's sake, the only data used for this analysis was those observations between 2018-01-01 and 2018-12-31.

```{r}
query <- "
WITH prep_data AS (
  SELECT
    *,
    CEIL((row_num - 1) / 6) + 1 as week
  FROM (
    SELECT
      *,
      ROW_NUMBER() OVER(ORDER BY date) as row_num
    FROM sp500_1950_2018 
    WHERE 
      date BETWEEN '2018-01-01' AND '2018-12-31'
    ORDER BY date
  )
)

SELECT
  date,
  open,
  high,
  AVG(open) OVER(ORDER BY date) as open_YTD_avg,
  AVG(high) OVER(ORDER BY date) as high_YTD_avg,
  AVG(open) OVER(PARTITION BY week ORDER BY week) as open_6_day_rolling_avg,
  AVG(high) OVER(PARTITION BY week ORDER BY week) as high_6_day_rolling_avg
FROM prep_data
WHERE 
  date BETWEEN '2018-01-01' AND '2018-12-31'
ORDER BY 1 
"
df <- sqldf(query)
head(df)
```
The `prep_data` CTE above shows how to use the `ROW_NUMBER` function to determine which 6-day "week" the observation falls into, and then the following query uses the CTE and window functions to determine the YTD and 6-day rolling average of both the opening and high daily price of of the S&P 500 stock index. 
