---
title: "Data 605 Discussion - Week 3"
author: "William Jasmine"
date: "2023-02-09"

output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(shiny)
```

Question comes from "A First Course in Linear Algebra" by Robert A. Beezer. 

# Question *T10*, Page 389

**Problem Description:**

A matrix $A$ is idempotent if $A^2 = A$. Show that the only possible eigenvalues of an idempotent matrix are $\lambda = 0, 1$. Then give an example of a matrix that is idempotent and has both of these two values as eigenvalues. 

---

**Proof:**

Consider a matrix $A$ and vector $x$ such that $\mathbf{x}$ is an eigenvector of $A$ with associated eigenvalue $\lambda$. In other words:

$$
\begin{equation}
A\mathbf{x} = \lambda\mathbf{x}
\end{equation}
$$

Given that $A^2 = A$, we also know that:

$$
\begin{equation}
A^2\mathbf{x} = \lambda\mathbf{x}
\end{equation}
$$

Based off the first equation and the properties of scalar multiplication, we can modify the expression $A^2\mathbf{x}$ as follows:

$$
\begin{equation}
A^2\mathbf{x} = A(Ax) = A(\lambda\mathbf{x}) = \lambda(A\mathbf{x}) = \lambda(\lambda\mathbf{x}) = \lambda^2\mathbf{x}
\end{equation}
$$
Thus combining the result of these above three equations we have that:

$$
\begin{align}
A\mathbf{x} &= \lambda\mathbf{x} \\
A\mathbf{x} &= \lambda^2\mathbf{x}
\end{align}
$$

The only way that this can be true is if $\lambda^2 = \lambda$. Thus:

$$
\begin{align}
\lambda^2 &= \lambda \\
\lambda^2 - \lambda &= 0 \\
\lambda(\lambda - 1) &= 0 \\
\end{align}
$$

The solutions to the above equation are $\lambda = 0$ and $\lambda = 1$. Thus, we conclude that is $A$ is an idempotent matrix, its only possible eigenvalues are 0 and 1. $\blacksquare$

---

**Example**:

The following matrix $A$ is idempotent:

$$
A = 
\begin{pmatrix}
3 & -6 \\
1 & -2 \\
\end{pmatrix}
$$

This is confirmed in the code cell below:

```{r}
A <- rbind(c(3, -6), c(1, -2))
A == A %*% A
```

To find the eigenvalues $\lambda$ of $A$, we can solve the following equation:

$$
\begin{align}
\text{det}
\begin{pmatrix}
3 - \lambda & -6 \\
1 & -2 - \lambda \\
\end{pmatrix} &= 0 \\
(3-\lambda)(-2-\lambda) + 6 &= \\
\lambda^2 -3\lambda + 2\lambda - 6 + 6 &= \\
\lambda^2 -\lambda = 0
\end{align}
$$

We see from the above that the eigenvalues of this idempotent matrix are indeed equal to 0 and 1. The math above is confirmed via the R `eigen` function:

```{r}
round(eigen(A)[[1]])

```
